Type Flattening Obfuscation

### 2020-01-28 paper/tool.tex
fix pointer arithmetic

\begin{document}
%
\title{Type Flattening Obfuscation}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Ta Thanh Dinh}
%
\authorrunning{Ta Thanh Dinh}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{
	\email{tathanhdinh@gmail.com}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
	Beside data and control flow, high-level types are also important in program analysis,
	indeed type reconstruction is an essential step in binary code decompilation.
	For the purpose of anti-decompilation, the paper introduces a novel obfuscation
	technique which makes types become harder to be reconstructed.

@@

	% We show that the techniques used for type reconstruction can be bypassed,
	% and we claim that the bypass is always possible given \emph{semantics gap} between
	% type systems of high-level and machine language.
	% \keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}
%
%
%
\section{Introduction}
An important step in binary code \emph{decompilation}~\cite{cifuentes_reverse_1994} is to
detect high-level objects (e.g. functions, variables) from low-level machine dependent
objects~\cite{balakrishnan_divine_2007} (e.g. registers, raw memory accesses, machine
instructions) before annotate them with types of the supposed original language. Because
most of information about high-level types is lost in compilation, the binary code
\emph{type reconstruction} requires special techniques, a survey up until 2015 can be
referenced in~\cite{caballero_type_2016}.

In the opposite direction, binary code \emph{obfuscation} is to make the decompilation (or in
general code analysis) harder. Since the data and control flow are principal elements for the
program analysis, current obfuscation techniques~\cite{collberg_surreptitious_2009} focus mostly
on hiding them. But to the best of our knowledge, there is still no explicit effort in hiding
high-level types. There would be several reasons for this lack of effort.

\emph{First}, type obfuscation can be obtained as side effects of data or control flow obfuscation.
Indeed, type reconstruction algorithms need both data and control flow to build type constraints,
if any of them is hidden
then the algorithms cannot work correctly. Or if the function boundary cannot be found
(because of anti-disassembing tricks, for example), then there are no inputs (i.e. high-level
objects) for type reconstruction. \emph{Second}, high-level types seem too coarse to be worthy
of being protected, in many cases just knowing certain values of the input which make the program
exploitable is enough. But knowledge about types expands attack surfaces because more analysis
can be proceeded, beside decompilation see the survey~\cite{caballero_type_2016} for a
more complete list.
% We exploit the \emph{semantics gap}

\section{Type flattening compiler}
We present \emph{uCc}, an open source C compiler that explicitly obfuscates high-level types.
Our effort is to attack the core of type reconstruction algorithms,
that is the data flow used to build type constraints, and the \emph{semantic gap} between types in the high-level language and their
representations in low-level machine code. For that purpose, \emph{uCc} uses extensively
obfuscation transformations based on Mixed Boolean Arithmetic expressions~\cite{eyrolles_obfuscation_2017,zhou_information_2007}.
To increase code diversity, it is a \emph{probabilistic} compiler: given the source
code, \emph{uCc} generates each time a different (but computationally equivalent) output machine code.

The compiler is implemented in Rust, uses \emph{cranelift}~\cite{noauthor_cranelift_nodate} as
backend. The source code and a short user guide are at (cite). We test outputs of
\emph{uCc} against decent decompilers, table~\ref{flattening} show experimental results

\begin{table}
	\begin{center}
		\caption{Functions and original types}
		\begin{tabular}{|l|l|}
			\hline
			Description & Function type \\
			\hline
			identity & \texttt{int id(int)} \\
			\hline
			division & \texttt{int div(short, char)} \\
			\hline
			modular & \texttt{int mod(int)} \\
			\hline
			increase pointer then dereference & \texttt{int inc\_deref(int*)} \\
			\hline
		\end{tabular}
	\end{center}
\end{table}

>>>>

\begin{document}
%
\title{Type Flattening Obfuscation}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Ta Thanh Dinh}
%
\authorrunning{Ta Thanh Dinh}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{
	\email{tathanhdinh@gmail.com}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
	Beside data and control flow, high-level types are also important in program analysis,
	indeed type recovery is an essential step in binary code decompilation.
	For the purpose of anti-decompilation, the paper introduces a novel obfuscation
	technique which makes types become harder to be reconstructed.

@@

	% We show that the techniques used for type reconstruction can be bypassed,
	% and we claim that the bypass is always possible given \emph{semantics gap} between
	% type systems of high-level and machine language.
	% \keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}
%
%
%
\section{Introduction}
An important step in binary code \emph{decompilation}~\cite{cifuentes_reverse_1994} is to
detect high-level objects (e.g. functions, variables) from low-level machine dependent
objects~\cite{balakrishnan_divine_2007} (e.g. registers, raw memory accesses, machine
instructions) before annotate them with types of the supposed original language. Because
most of information about high-level types is lost in compilation, the binary code
\emph{type recovery} requires special techniques, a survey up until 2015 can be
referenced in~\cite{caballero_type_2016}.

In the opposite direction, binary code \emph{obfuscation} is to make the decompilation (or in
general code analysis) harder. Since the data and control flow are principal elements for the
program analysis, current obfuscation techniques~\cite{collberg_surreptitious_2009} focus mostly
on them. But to the best of our knowledge, there is still no explicit effort in obfuscating
high-level types. There would be several reasons for this lack of effort.

\emph{First}, type obfuscation is unsurprisingly a side effect of data or control flow obfuscation.
Indeed, type recovery algorithms need both data and control flow to build type constraints,
if any of them is hidden
then the algorithms cannot work correctly. Or if the function boundary is not found
(because of anti-disassembing tricks, for example), then high-level objects cannot be recognized.
\emph{Second}, high-level types seem too coarse to be worthy
of being protected, in many cases just knowing certain values of the input which make the program
exploitable is enough. But knowledge about types expands attack surfaces because more analysis
can be proceeded, beside decompilation see the survey~\cite{caballero_type_2016} for a
more complete list.
% We exploit the \emph{semantics gap}

\section{Type flattening compiler}
We present \emph{uCc}, an open source C compiler that explicitly obfuscates high-level types.
The compiler is implemented in Rust, its source code and a short user guide are given at (cite).
Our effort is to attack the core of type recovery algorithms,
these are the data flow used to build type constraints, and the \emph{semantic gap} between types in the high-level language and their
representations in the low-level machine code. Beside custom tricks, \emph{uCc} uses extensively
obfuscation transformations based on Mixed Boolean Arithmetic expressions~\cite{eyrolles_obfuscation_2017,zhou_information_2007}.
To make some code diversity, the compiler is \emph{probabilistic}: given a source
code, \emph{uCc} generates each time a different (but computationally equivalent) output machine code.

We test outputs of \emph{uCc} against some decompilers, table~\ref{description} describes briefly
tested functions and their types in source code, table~\ref{flattening} shows results of using
decompilers to decompile binaries generated by \emph{uCc}. The test suite is also available at
the repository.

\begin{table}
	\begin{center}
		\caption{Functions and original types}\label{description}
		\begin{tabular}{|l|l|}
			\hline
			Description & Function type \\
			\hline
			identity & \texttt{int id(int)} \\
			\hline
			division & \texttt{int div(short, char)} \\
			\hline
			modular & \texttt{int mod(int)} \\
			\hline
			increase pointer then dereference & \texttt{int inc\_deref(int*)} \\
			\hline
		\end{tabular}
	\end{center}
\end{table}
